---
title: "Advanced Next-Generation Workflows in PubChemR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Next-Generation Workflows in PubChemR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette covers advanced features that are not the focus of the legacy
`get_*`/PUG REST/PUG View tutorials. It focuses on resilient, typed, and
high-throughput `pc_*` workflows:

1. Typed result contracts and error handling.
2. Unified request layer and endpoint wrappers.
3. Caching, offline replay, and runtime policy control.
4. Async listkey workflows.
5. Checkpointed batching and resume patterns.
6. Benchmark harness and threshold-gated performance checks.
7. Analysis-layer helpers, sparse matrices, and model export.
8. Optional cheminformatics bridges and lifecycle policy.

```{r setup}
library(PubChemR)
library(dplyr)
library(tibble)
```

## 1) Typed result contract

The `pc_*` layer returns structured objects (`PubChemResult`,
`PubChemBatchResult`, `PubChemBenchmarkReport`) with explicit
`success`/`error` fields.

```{r}
# Deterministic local payload -> typed successful result:
ok_text <- '{"PropertyTable":{"Properties":[{"CID":2244,"MolecularWeight":180.16}]}}'
res <- pc_response(
  ok_text,
  request = list(
    domain = "compound",
    namespace = "cid",
    identifier = 2244,
    operation = "property/MolecularWeight",
    output = "JSON"
  )
)

res$success
as_tibble(res)
```

You can also normalize raw payload text into the same typed structure:

```{r}
fault_text <- '{"Fault":{"Code":"PUGREST.BadRequest","Message":"Invalid request"}}'
fault_res <- pc_response(fault_text, request = list(url = "https://example.org"))
fault_res$success
fault_res$error$code
```

## 2) Runtime policy and transport controls

Use `pc_profile()` for common policies, then override with `pc_config()`.

```{r}
pc_profile("default")
cfg <- pc_config(
  rate_limit = 5,
  timeout = 60,
  retries = 3,
  pause_base = 1,
  pause_cap = 8,
  cache_ttl = 24 * 60 * 60
)
cfg
```

Input validation is strict for chunk sizes, worker counts, and numeric controls.

```{r}
try(pc_config(rate_limit = 0), silent = TRUE)
try(pc_request(cache_ttl = Inf, offline = TRUE), silent = TRUE)
```

## 3) Unified request layer and typed wrappers

`pc_request()` is the generic entry point. Wrappers provide typed intent.

```{r eval=FALSE}
# Generic request
r0 <- pc_request(
  domain = "compound",
  namespace = "cid",
  identifier = 2244,
  operation = "property/MolecularWeight,XLogP",
  output = "JSON"
)

# Domain-oriented wrappers
cmp <- pc_compound(identifier = 2244, namespace = "cid", operation = "record")
sub <- pc_substance(identifier = 2244, namespace = "cid", operation = "record")
asy <- pc_assay(identifier = 1000, namespace = "aid", operation = "summary")
prp <- pc_property(identifier = c(2244, 3672), properties = c("MolecularWeight", "XLogP"))
idm <- pc_identifier_map(identifier = "aspirin", namespace = "name", to = "cids")

as_tibble(prp)
as_tibble(idm)
```

## 4) Caching and offline replay

PubChemR supports memory+disk caching, diagnostics, and cache-only replay mode.

```{r eval=FALSE}
pc_cache_clear()

first <- pc_property(
  identifier = 2244,
  properties = c("MolecularWeight", "XLogP"),
  namespace = "cid",
  cache = TRUE
)

second <- pc_property(
  identifier = 2244,
  properties = c("MolecularWeight", "XLogP"),
  namespace = "cid",
  cache = TRUE
)

second$from_cache
pc_cache_info()
```

Offline replay mode is useful for deterministic pipelines:

```{r}
offline_miss <- pc_request(
  domain = "compound",
  namespace = "cid",
  identifier = 2244,
  output = "JSON",
  cache = TRUE,
  offline = TRUE
)

offline_miss$success
offline_miss$error$code
```

## 5) Async listkey workflows (`pc_submit` -> `pc_poll` -> `pc_collect`)

For asynchronous queries, use:

```{r eval=FALSE}
q <- pc_submit(
  domain = "compound",
  namespace = "name",
  identifier = "aspirin",
  searchtype = "similarity",
  operation = "cids",
  output = "JSON",
  options = list(Threshold = 95, MaxRecords = 200)
)

# Option A: explicit polling
final_a <- pc_poll(q, interval = 1.5, max_attempts = 20)

# Option B: convenience collector
final_b <- pc_collect(q, interval = 1.5, max_attempts = 20)

as_tibble(final_b)
```

## 6) Checkpointed batching and resume

`pc_batch()` supports chunking, optional parallelism, and resumable checkpoints.

```{r eval=FALSE}
ids <- c("aspirin", "ibuprofen", "caffeine", "acetaminophen", "naproxen")
cp_dir <- file.path(tempdir(), "pc_batch_checkpoint")

batch <- pc_batch(
  ids = ids,
  fn = function(chunk_ids) {
    pc_identifier_map(
      identifier = chunk_ids,
      namespace = "name",
      to = "cids",
      domain = "compound",
      cache = TRUE
    )
  },
  chunk_size = 2,
  checkpoint_dir = cp_dir,
  checkpoint_id = "name_to_cid_demo"
)

batch
as_tibble(batch)
```

Resume from existing checkpoint state:

```{r eval=FALSE}
resumed <- pc_resume_batch(
  fn = function(chunk_ids) {
    pc_identifier_map(
      identifier = chunk_ids,
      namespace = "name",
      to = "cids",
      domain = "compound",
      cache = TRUE
    )
  },
  checkpoint_dir = cp_dir,
  checkpoint_id = "name_to_cid_demo"
)

resumed$checkpoint$resumed
as_tibble(resumed)
```

## 7) Benchmarking and threshold gates

`pc_benchmark()` compares chunk and parallel scenarios.  
`pc_benchmark_harness()` scales this into scenario gates (10/1k/100k by default).

```{r eval=FALSE}
probe <- function(ids) {
  pc_request(
    domain = "compound",
    namespace = "cid",
    identifier = ids,
    operation = "property/MolecularWeight",
    output = "JSON",
    cache = FALSE
  )
}

bm <- pc_benchmark(
  ids = rep(2244, 200),
  fn = probe,
  chunk_sizes = c(25, 50, 100),
  parallel_options = c(FALSE)
)
bm
```

```{r eval=FALSE}
thresholds <- list(
  elapsed_sec = c(`10` = 30, `1000` = 300, `100000` = 3600),
  failed_chunk_ratio = c(`10` = 0, `1000` = 0.01, `100000` = 0.05)
)

h <- pc_benchmark_harness(
  fn = probe,
  ids = rep(2244, 100000),
  scenario_sizes = c(10L, 1000L, 100000L),
  chunk_sizes = c(100L, 1000L),
  thresholds = thresholds,
  report_path = file.path(tempdir(), "pubchemr-benchmark.md"),
  report_format = "markdown"
)

h$summary
```

## 8) Analysis-layer helpers for assay modeling

### 8.1 Similarity retrieval and long assay activity table

```{r eval=FALSE}
sim <- pc_similarity_search(
  identifier = "CC(=O)OC1=CC=CC=C1C(=O)O",
  namespace = "smiles",
  threshold = 90,
  max_records = 200,
  cache = TRUE
)

sim_tbl <- as_tibble(sim) %>%
  filter(!is.na(CID)) %>%
  mutate(CID = as.character(CID)) %>%
  distinct(CID)

assay_long <- pc_assay_activity_long(
  identifier = sim_tbl$CID,
  namespace = "cid",
  chunk_size = 25,
  cache = TRUE
)
```

`pc_assay_activity_long()` now fails explicitly if any chunk fails, instead of
silently dropping failed chunks.

### 8.2 Activity matrix (dense or sparse)

```{r eval=FALSE}
activity_dense <- pc_activity_matrix(
  assay_long,
  cid_col = "CID",
  aid_col = "AID",
  outcome_col = "ActivityOutcome",
  aggregate = "max",
  fill = NA_real_,
  output = "tibble"
)

activity_sparse <- pc_activity_matrix(
  assay_long,
  cid_col = "CID",
  aid_col = "AID",
  outcome_col = "ActivityOutcome",
  aggregate = "max",
  fill = NA_real_,
  output = "sparse"
)

activity_sparse
```

### 8.3 Feature table, matrix conversion, and export

```{r eval=FALSE}
feat <- pc_feature_table(
  identifier = unique(assay_long$CID),
  properties = c("MolecularWeight", "XLogP", "TPSA"),
  namespace = "cid",
  cache = TRUE
)

joined <- pc_cross_domain_join(
  compounds = feat,
  assays = assay_long %>% select(CID, AID, ActivityOutcome)
)

mm <- pc_model_matrix(
  x = feat,
  outcome = NULL,
  id_cols = c("CID"),
  na_fill = 0,
  scale = TRUE
)

pc_export_model_data(mm, path = file.path(tempdir(), "pubchemr_model.csv"), format = "csv")
pc_export_model_data(mm, path = file.path(tempdir(), "pubchemr_model.rds"), format = "rds")
```

## 9) Optional ecosystem bridges

These helpers require optional packages:

- `pc_to_rcdk()` requires `rcdk`
- `pc_to_chemminer()` requires `ChemmineR`

```{r eval=FALSE}
# Example assumes a table with CanonicalSMILES column:
mols <- pc_to_rcdk(feat, smiles_col = "CanonicalSMILES", id_col = "CID")
sdf  <- pc_to_chemminer(feat, smiles_col = "CanonicalSMILES")
```

## 10) Targeted PUG View helper for bio test sections

Besides generic `get_pug_view()`, `get_biological_test_results()` directly
extracts the `"Biological Test Results"` section.

```{r eval=FALSE}
bio <- get_biological_test_results(
  identifier = "2244",
  domain = "compound",
  heading = "Biological Test Results",
  .all = TRUE
)

bio
```

## 11) Lifecycle policy (legacy vs nextgen)

```{r}
pc_lifecycle_policy()
```

## 12) Recommended production pattern

For robust pipelines:

1. Set policy with `pc_profile()`/`pc_config()`.
2. Enable cache and inspect with `pc_cache_info()`.
3. Use typed wrappers (`pc_property()`, `pc_identifier_map()`, etc.).
4. Run heavy jobs through `pc_batch()` with checkpoints.
5. Use `pc_resume_batch()` for interruptions.
6. Gate performance with `pc_benchmark_harness()`.
7. Export model-ready artifacts with `pc_export_model_data()`.
