---
title: "Enhancing Chemical Data Access with PubChemR: A Guide to Utilizing PUG View Service"
author: "Selcuk Korkmaz, Dincer Goksuluk, Bilge Eren Korkmaz"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true

vignette: >
  %\VignetteIndexEntry{Enhancing Chemical Data Access with PubChemR: A Guide to Utilizing PUG View Service}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache.path = "vignettes/cache/"
)
library(PubChemR)
```

```{css, echo=FALSE}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

```{css, echo=FALSE}
.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
  background-color: inherit;
}
```

## 1. Introduction

The PubChemR package features the essential function get_pug_view, specifically developed to provide access to comprehensive summary reports and additional information not typically found in the primary PubChem Substance, Compound, or BioAssay records. Utilizing the PUG View service, a REST-style web service of PubChem, get_pug_view is primarily utilized for generating detailed summaries for the PubChem database record web pages. It can also function as a standalone programmatic web service.

PUG View is expertly designed to provide complete summary reports on individual PubChem records, offering a different approach compared to the PUG REST service, which delivers smaller bits of information about one or more PubChem records. The get_pug_view function sends requests to the PubChem PUG View API, allowing users to retrieve various types of data for a given identifier, including annotations and QR codes. It supports multiple output formats such as JSON and SVG, making it a versatile tool for users who need comprehensive information from PubChem's database. This vignette aims to elucidate the structure and usage of the PUG View service, offering illustrative use cases to aid new users in understanding its operation and constructing effective requests.

## 2. Accessing PUG View with `get_pug_view`

PUG View offers a versatile approach to access structured information from PubChem's extensive database. Here's how to leverage the get_pug_view function in R for accessing various data formats and specific record summaries:

### 2.1. Data Formats

**Multiple Formats:** PUG View supports JSON, JSONP, XML, ASNT, and ASNB formats. For example, to retrieve compound data in JSON format:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "data", identifier = "1234", domain = "compound", output = "JSON")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

**XML Schema:** An XML schema is available for users requiring a structured format. This can be accessed via https://pubchem.ncbi.nlm.nih.gov/pug_view/pug_view.xsd

### 2.2. Record Summaries and Full Records

**Index vs Full Data:** Choose between an index (a summary or table of contents) and full data retrieval. For example, to get a compound index:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "index", identifier = "1234", domain = "compound", output = "XML")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

**Specific Headings:** Retrieve data under specific headings for more targeted information. For instance, to get experimental properties:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "data", identifier = "2244", domain = "compound", output = "JSON", heading = "Experimental Properties")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

### 2.3. Accessing Different Record Types

**Compounds, Substances, and BioAssays:** Access these records using their respective identifiers (CID, SID, AID). For example, to access a BioAssay record:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "data", identifier = "1", domain = "assay", output = "JSON")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

**Patents, Genes, Proteins, Pathways, Taxonomies, Cell Lines, Elements:** Each of these can be accessed using their specific identifiers or names. For instance, to retrieve information on a specific gene:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "data", identifier = "1", domain = "gene", output = "JSON")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

### 2.4. Annotations

**Access by Heading:** Retrieve specific types of information across PubChem's databases. For example, to get all experimental viscosity measurements:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "annotations", identifier = "Viscosity",   domain = "heading", output = "JSON")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

Or equivalently (useful if the heading contains special characters not compatible with URL syntax):

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "annotations", identifier = NULL,   domain = "heading", output = "JSON", heading = "Viscosity")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```


**Specify Heading Type:** For headings referring to different record types, specify the heading type. For example, boiling point for compounds:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "annotations", identifier = "Boiling Point", domain = "heading", output = "XML", headingType = "Compound")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

**Pagination:** Access different pages of data if the total exceeds one page. For example, the 10th page of CAS annotations:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "annotations", identifier = "CAS",domain = "heading", output = "JSON", page = "10")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

### 2.5. Source Categories

**Depositors and SIDs:** List all depositors for a given compound, categorized by source type:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "categories", identifier = "1234", domain = "compound", output = "JSON")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

### 2.6. Neighbors

**Similar Compounds:** Get a list of compounds with similar structures and associated information:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "neighbors", identifier = "1234", domain = "compound", output = "JSON")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

### 2.7. Literature

**PubMed URLs:** Retrieve literature associated with a compound, organized by subheading:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "literature", identifier = "1234", domain = "compound", output = "JSON")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

### 2.8. Biologics

**Biologic Images:** Display biologic images associated with compounds:

```{r, class.output="scroll-100", cache=TRUE, eval=FALSE}
get_pug_view(annotation = "image", identifier = "243577", domain = "biologic", output = "SVG")
```

### 2.9. QR Codes

**Safety and Hazard Labelling:** Generate QR codes linking to LCSS pages for compounds:

```{r, class.output="scroll-100", cache=TRUE, eval=FALSE}
get_pug_view(annotation = "qr", identifier = "1234", domain = "compound", output = "SVG", qrSize = "short")
```

### 2.10. Linkout

**NCBI LinkOut Records:** List all LinkOut records for a substance, compound, or assay:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "linkout", identifier = "1234", domain = "compound", output = "JSON")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

### 2.11. PDB/MMDB Structures

**3D Protein Structures:** List 3D protein structures associated with a compound:

```{r, class.output="scroll-100", cache=TRUE}
result <- get_pug_view(annotation = "structure", identifier = "2244", domain = "compound", output = "JSON")
```

```{r, class.output="scroll-100", cache=TRUE}
result
```

### 2.12. Annotation Attachments

**Specialized Retrieval:** Access attachments like spectral images associated with records:

```{r, class.output="scroll-100", cache=TRUE, eval=FALSE}
get_pug_view(annotation = "data", identifier = "236678_1", domain = "key")
```


