---
title: "Phase 4 Analysis Layer: Similarity to Model-Ready Export"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phase 4 Analysis Layer: Similarity to Model-Ready Export}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette demonstrates the first true analysis-layer workflow in PubChemR:

1. Similarity retrieval around a seed compound.
2. Assay activity extraction in long format.
3. Activity matrix construction.
4. Feature retrieval and model-matrix assembly.
5. Model-ready export.

```{r setup}
library(PubChemR)
library(dplyr)
library(tibble)
```

## 1) Similar compounds from a seed structure

```{r eval=FALSE}
# Aspirin SMILES as seed
seed_smiles <- "CC(=O)OC1=CC=CC=C1C(=O)O"

sim <- pc_similarity_search(
  identifier = seed_smiles,
  namespace = "smiles",
  threshold = 90,
  max_records = 200,
  cache = TRUE
)

sim_tbl <- as_tibble(sim) %>%
  filter(!is.na(CID)) %>%
  mutate(CID = as.character(CID)) %>%
  distinct(CID)

sim_tbl
```

## 2) Fetch assay activity in long format

```{r eval=FALSE}
assay_long <- pc_assay_activity_long(
  identifier = sim_tbl$CID,
  namespace = "cid",
  chunk_size = 25,
  cache = TRUE
)

assay_long %>%
  select(CID, AID, ActivityOutcome, ActivityValue_uM) %>%
  head()
```

## 3) Build activity matrix (compound x assay)

```{r eval=FALSE}
activity_mat <- pc_activity_matrix(
  assay_long,
  cid_col = "CID",
  aid_col = "AID",
  outcome_col = "ActivityOutcome",
  aggregate = "max",
  fill = NA_real_
)

activity_mat
```

## 4) Add chemical features and prepare model matrix

```{r eval=FALSE}
feature_tbl <- pc_feature_table(
  identifier = sim_tbl$CID,
  properties = c(
    "MolecularWeight",
    "XLogP",
    "TPSA",
    "HBondDonorCount",
    "HBondAcceptorCount"
  ),
  namespace = "cid",
  cache = TRUE
) %>%
  mutate(CID = as.character(CID))

model_tbl <- feature_tbl %>%
  left_join(activity_mat, by = "CID")

mm <- pc_model_matrix(
  x = model_tbl,
  id_cols = "CID",
  na_fill = 0,
  scale = TRUE
)

mm
```

## 5) Export model-ready artifact

```{r eval=FALSE}
out_csv <- file.path(tempdir(), "phase4_similarity_activity_model.csv")
out_rds <- file.path(tempdir(), "phase4_similarity_activity_model.rds")

pc_export_model_data(mm, path = out_csv, format = "csv")
pc_export_model_data(mm, path = out_rds, format = "rds")

out_csv
out_rds
```

This workflow is intended for reproducible discovery pipelines where
structure-driven retrieval, bioactivity reshaping, and model preparation are
performed within one package.
