---
title: "PubChemR Workflow: Retrieval to Modeling Table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PubChemR Workflow: Retrieval to Modeling Table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates a reproducible pattern for working with PubChem data:

1. Identify compounds.
2. Retrieve selected properties.
3. Build a model-ready table.
4. Convert to a modeling matrix.

```{r setup}
library(PubChemR)
library(dplyr)
```

## 1) Resolve CIDs from names

```{r eval=FALSE}
ids <- get_cids(c("aspirin", "ibuprofen", "caffeine"), namespace = "name")
CIDs(ids)
```

## 2) Retrieve a compact property panel

```{r eval=FALSE}
props <- get_properties(
  properties = c("MolecularWeight", "MolecularFormula", "XLogP", "TPSA"),
  identifier = c("aspirin", "ibuprofen", "caffeine"),
  namespace = "name"
)

prop_tbl <- retrieve(props, .combine.all = TRUE, .to.data.frame = TRUE)
prop_tbl
```

## 3) Prepare a modeling-ready table

```{r eval=FALSE}
model_tbl <- prop_tbl %>%
  mutate(
    MolecularWeight = as.numeric(MolecularWeight),
    XLogP = as.numeric(XLogP),
    TPSA = as.numeric(TPSA)
  )

model_tbl
```

## 4) Convert to model matrix form

```{r eval=FALSE}
mm <- pc_model_matrix(
  model_tbl,
  id_cols = c("CID"),
  na_fill = 0
)

mm
```

For high-throughput workflows, combine this pattern with deterministic identifier sets
and saved intermediate outputs for reproducibility.
